---
title: "Vancouver Tree Height Geography"
author: "DSCI 522, Group 33"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Vancouver Tree Height Geography

## Summary

Our group was interested in answering the question, "**Do tree heights vary significantly from neighborhood to neighborhood in Vancouver?**"

We aim to analyze the relationship between tree height distribution and neighborhoods. Specifically, we focus on tree height ranges and their counts in various neighborhoods to determine if tree height is influenced by location.

## Introduction

Urban trees are essential to creating livable cities, offering ecological, aesthetic, and health benefits. They improve air quality, provide shade, support biodiversity, and enhance the overall urban environment. In Vancouver, street trees play a vital role in the city’s commitment to sustainability. However, the characteristics of these trees, such as their heights, can vary significantly across neighborhoods due to factors like local urban planning, soil quality, and maintenance practices. Understanding these patterns is key to equitable urban forestry management and informed decision-making.

This project explores the relationship between **tree height distribution** and **neighborhoods** in Vancouver. Using data from the [City of Vancouver Open Data Portal](https://opendata.vancouver.ca/explore/dataset/street-trees/information/?disjunctive.species_name&disjunctive.common_name&disjunctive.on_street&disjunctive.neighbourhood_name), we seek to answer the question: **Do tree heights vary significantly between neighborhoods?**

To address this, we analyze the dataset through a combination of:

1.  **Exploratory Data Analysis (EDA):** We use contingency tables and visualizations (heatmaps) to identify patterns in tree height distributions across neighborhoods.

2.  **Statistical Testing:** A Chi-squared test of independence is performed to determine if the observed variations in tree height distributions are statistically significant.

By uncovering these patterns, this analysis contributes to urban forestry strategies that aim to distribute greenery benefits equitably across neighborhoods. The findings could help guide future decisions in tree planting, maintenance, and sustainable urban planning.

## Methods and Results

### Loading Required Packages

```{r}
library(tidyverse)
library(janitor)
library(ggplot2)
library(knitr)
```

### Loading the Data

```{r}
trees <- read_csv2("data/street-trees.csv")
head(trees)
```

### Cleaning the Data

We are only interested in neighbourhood names and tree heights, so we should make sure there are no missing values in these columns

```{r}
str(trees) # maybe we don't use this
# look at NEIGHBOURHOOD_NAME, HEIGHT_RANGE, HEIGHT_RANGE_ID
```

### Exploratory Data Analysis

The columns of interest are:

1.  `NEIGHBOURHOOD_NAME` (explanatory/treatment variable) - A string representing the neighbourhood the tree is in

2.  Tree height data (the outcome/response variable) is represented in two columns, in different formats:

    1.  `HEIGHT_RANGE` - a string representing tree heights (categorical levels) in buckets of 10ft, e.g. `0' - 10'`, `10' - 20'`, etc.

    2.  `HEIGHT_RANGE_ID` - a numeric column (integers) with levels corresponding to the above strings

First, we should examine the levels of the two available versions of the response categorical variable (height), because we may be able to choose one that will simplify our subsequent analysis.

Based on the data, it looks like there should be a one-to-one correspondence between levels of `HEIGHT_RANGE` and `HEIGHT_RANGE_ID`, but we should confirm this. To do this, we can look at all unique combinations of the two variables. If they properly correspond (there are no issues with the data), we should see only one row for each. We will print the results using `kable()` (from the `knitr` package).

```{r}
unique_combinations <- trees |> 
  distinct(HEIGHT_RANGE, HEIGHT_RANGE_ID) |> 
  arrange(HEIGHT_RANGE_ID)

unique_combinations |>
  kable(caption = "Mapping between tree height category names and their IDs.")
```

It looks like there is a proper correspondence between the levels of these two variables. Since the levels of `HEIGHT_RANGE_ID` are simpler and shorter, we will use this variable. This should make our plots easier to display.

Now we can select only the treatment and response variable columns and convert them to type `factor`, which will allow us to build a contingency table.

```{r}
trees_subset <- trees |> 
  select(NEIGHBOURHOOD_NAME, HEIGHT_RANGE_ID) |> 
  mutate(across(everything(), as.factor))

head(trees_subset)
```

The new `trees_subset` dataframe contains one row per tree, with only the columns `NEIGHBOURHOOD_NAME` and `HEIGHT_RANGE_ID`. We can use this along with the `tabyl()` function from the `janitor` package to generate a contingency table. We will then print the contingency table using `kable()`.

```{r}
cont_table <- trees_subset |> 
  tabyl(NEIGHBOURHOOD_NAME, HEIGHT_RANGE_ID)

cont_table |> 
  kable(caption = "Contingency table showing counts of trees in various levels of the tree height variable per levels of the neighbourhood variable. For a mapping of tree height category names, refer to Table 1.")
```

In order to better understand patterns in this data, we can visualize the above contingency table as a two-dimensional histogram (a.k.a heat map):

```{r}
heatmap_data <- cont_table |> 
  pivot_longer(
    cols = -NEIGHBOURHOOD_NAME,
    names_to = "HEIGHT_RANGE_ID",
    values_to = "Tree_Count"
  ) |> 
  mutate(
    HEIGHT_RANGE_ID = factor(HEIGHT_RANGE_ID,
                             levels = as.character(0:10))
  )

ggplot(heatmap_data, aes(x = HEIGHT_RANGE_ID,
                         y = NEIGHBOURHOOD_NAME,
                         fill = Tree_Count)) +
  geom_tile() +
  scale_fill_gradient(low = "white",
                      high = "blue") +
  labs(
    title = "Figure 1: Tree Height Distribution by Neighbourhood",
    x = "Height Category ID",
    y = "Neighbourhood",
    fill = "Tree Count"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

From the plot, it looks like there may be some differences in tree heights across neighbourhoods. In particular:

-   RENFREW-COLLINGWOOD has, by a large margin, the most trees in the 10’ - 20’ height category (ID 1)

-   VICTORIA-FRASERVIEW, SUNSET, RENFREW-COLLINGWOOD, KENSINGTON-CEDAR COTTAGE, and HASTINGS-SUNRISE seem to have more trees in the lower height categories (1-4) than other neighbourhoods.

-   SHAUGHNESSY and DOWNTOWN seem to have the tallest trees.

### Statistical Test

Although there are some visually identifiable patterns, we can only determine whether there are statistically significant differences in tree heights per neighbourhood using an appropriate test.

#### Choosing a Test and Significance Level

Because we are dealing with two categorical variables, each with multiple levels, a Chi-squared test of Independence/Homogeneity is appropriate.

The Chi-squared test makes the following assumptions:

1.  The observations are independent.
2.  The expected counts are large enough (greater than 5 is typical).

First, the height measurement of a particular tree does not depend on that of another, so we can assume independence. Second, almost every intersection of the contingency table has counts much larger than 5, with many in the hundreds or thousands.

Because neither of the test's two key assumptions appear to be violated, it is safe to proceed with a Chi-squared test of independence.

Finally, we will choose the standard significance level of $\alpha = 0.05$ as our threshold for determining statistical significance.

#### Performing a Chi-squared Test of Independence

We will perform the test, store the results in `chisq_results`, and display them.

```{r}
chisq_results <- chisq.test(cont_table, correct = FALSE)

chisq_results
```

The Chi-squared test yielded a statistically significant result, with a test statistic of $X^2 = 15339$ and $p < 2.2 \times 10^{-16}$, which is less than our predefined significance level of 0.05.

## Discussion

sd

## References

-   City of Vancouver Open Data Portal: *Street Trees Dataset*. Available at: <https://opendata.vancouver.ca/explore/dataset/street-trees/information/?disjunctive.species_name&disjunctive.common_name&disjunctive.on_street&disjunctive.neighbourhood_name>

-   Wickham, H., & Grolemund, G. (2016). *R for Data Science: Import, Tidy, Transform, Visualize, and Model Data*. O'Reilly Media.

-   Janitor R Package Documentation. Available at: <https://cran.r-project.org/web/packages/janitor/janitor.pdf>

-   ggplot2 R Package Documentation. Available at: <https://ggplot2.tidyverse.org/>

-   Vancouver Urban Forestry Strategy. (2018). City of Vancouver. Available at: <https://vancouver.ca/parks-recreation-culture/urban-forestry-strategy.aspx>
