---
title: "Vancouver Tree Height Geography"
author: "DSCI 522, Group 33"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Vancouver Tree Height Geography

## Summary

Our group was interested in answering the question, "**Do tree heights vary significantly from neighborhood to neighborhood in Vancouver?**"

We aim to analyze the relationship between tree height distribution and neighborhoods. Specifically, we focus on tree height ranges and their counts in various neighborhoods to determine if tree height is influenced by location.

## Introduction

Urban trees are essential to creating livable cities, offering ecological, aesthetic, and health benefits. They improve air quality, provide shade, support biodiversity, and enhance the overall urban environment. In Vancouver, street trees play a vital role in the city’s commitment to sustainability. However, the characteristics of these trees, such as their heights, can vary significantly across neighborhoods due to factors like local urban planning, soil quality, and maintenance practices. Understanding these patterns is key to equitable urban forestry management and informed decision-making.

This project explores the relationship between **tree height distribution** and **neighborhoods** in Vancouver. Using data from the [City of Vancouver Open Data Portal](https://opendata.vancouver.ca/explore/dataset/street-trees/information/?disjunctive.species_name&disjunctive.common_name&disjunctive.on_street&disjunctive.neighbourhood_name)

To address this, we analyze the dataset through a combination of:

1.  **Exploratory Data Analysis (EDA):** We use contingency tables and visualizations (heatmaps) to identify patterns in tree height distributions across neighborhoods.

2.  **Statistical Testing:** A Chi-squared test of independence is performed to determine if the observed variations in tree height distributions are statistically significant.

By uncovering these patterns, this analysis contributes to urban forestry strategies that aim to distribute greenery benefits equitably across neighborhoods in Vancouver. The findings could help guide future decisions in tree planting, maintenance, and sustainable urban planning.

## Methods and Results

### Loading Required Packages

```{r}
library(tidyverse)
library(janitor)
library(ggplot2)
library(knitr)
```

### Loading the Data

```{r}
trees <- read_csv2("data/street-trees.csv")
head(trees)
```

### Cleaning the Data

In our analysis, we are primarily interested in the NEIGHBOURHOOD_NAME, HEIGHT_RANGE, and HEIGHT_RANGE_ID columns, so it is crucial to ensure that there are no missing values in these columns to maintain the accuracy and reliability of our statistical results. Upon checking, we confirmed that none of these three columns contain any missing data.   

Additionally, we examined the entire dataset for duplicate records using the unique identifier TREE_ID and confirmed that there are no duplicate rows. This step ensures that no records are inadvertently counted multiple times.

```{r}
# str(trees) # maybe we don't use this
# Check for missing data in NEIGHBOURHOOD_NAME, HEIGHT_RANGE, HEIGHT_RANGE_ID columns
missing_data_check <- trees |>
  summarise(
    NEIGHBOURHOOD_NAME_missing = sum(is.na(NEIGHBOURHOOD_NAME)),
    HEIGHT_RANGE_missing = sum(is.na(HEIGHT_RANGE)),
    HEIGHT_RANGE_ID_missing = sum(is.na(HEIGHT_RANGE_ID))
  )
print(missing_data_check) # no missing data in the columns of interest

# Check for duplicates in data set
duplicate_count <- trees |>
  select(TREE_ID) |>
  duplicated() |>
  sum()
print(paste("Number of duplicate records:", duplicate_count)) # no duplicated records
```

### Exploratory Data Analysis

The columns of interest are:

1\. `NEIGHBOURHOOD_NAME` (explanatory/treatment variable) - A string representing the neighbourhood the tree is in

2\. Tree height data (the outcome/response variable) is represented in two columns, in two different formats:

\- `HEIGHT_RANGE` - a string representing tree heights (categorical levels) in buckets of 10ft, e.g. `0' - 10'`, `10' - 20'`, etc.

\- `HEIGHT_RANGE_ID` - a numeric column (integers) representing levels of the above strings

First, we should examine the levels of the two available versions of the treatment categorical variable we could choose one that will simplify our subsequent analysis.

Based on the data, it looks like there should be a one-to-one correspondence between levels of `HEIGHT_RANGE` and `HEIGHT_RANGE_ID`, but we should confirm this. To do this, we can look at all unique combinations of the two variables. If they properly correspond (there are no issues with the data), we should see only one row for each.

```{r}
unique_combinations <- trees |> 
  distinct(HEIGHT_RANGE, HEIGHT_RANGE_ID) |> 
  arrange(HEIGHT_RANGE_ID)

unique_combinations |> kable()
#TODO: add figure titles to this (in the kable code) and others

```

It looks like there is proper correspondence between levels of these two variables. Since the levels of `HEIGHT_RANGE_ID` are simpler and shorter, we will use this variable, which should make our plots easier to display.

Now we can select only the treatment and response variable columns and convert them to type `factor`. Then, we can display the con

```{r}
trees_subset <- trees |> 
  select(NEIGHBOURHOOD_NAME, HEIGHT_RANGE_ID) |> 
  mutate(across(everything(), as.factor))

trees_subset
```

Now, the `trees` tibble contains a row per tree, with only the columns `HEIGHT_RANGE` and `HEIGHT_RANGE_ID`.

Generate a contingency table and print it using `kable()`

NOTE TO SELF: add figure captions and for the one below, refer to the tibble above

```{r}
cont_table <- trees_subset |> 
  tabyl(NEIGHBOURHOOD_NAME, HEIGHT_RANGE_ID)

cont_table |> 
  kable()
```

In order to better understand patterns in this data, we can visualize the above contingency table as a 2-dimensional histogram (a.k.a heat map):

```{r}
heatmap_data <- cont_table |> 
  pivot_longer(
    cols = -NEIGHBOURHOOD_NAME,
    names_to = "HEIGHT_RANGE_ID",
    values_to = "Tree_Count"
  ) |> 
  mutate(
    HEIGHT_RANGE_ID = factor(HEIGHT_RANGE_ID, levels = as.character(0:10))
  )

ggplot(heatmap_data, aes(x = HEIGHT_RANGE_ID, y = NEIGHBOURHOOD_NAME, fill = Tree_Count)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "blue") +
  labs(
    title = "Tree Distribution by Neighbourhood and Height Range",
    x = "Height Range",
    y = "Neighbourhood",
    fill = "Tree Count"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

From the plot, it looks like there may be some significant differences in tree heights across neighbourhoods. In particular:

-   sd

-   sd

### Statistical Test

is it statistically significant?

#### Choosing the Appropriate Test

Because we are dealing with two categorical variables each with multiple levels, a Chi-squared test of Independence/Homogeneity is appropriate

The Chi-squared test makes the following assumptions:

1.  The observations are independent.
2.  The expected counts are sufficiently large (greater than 5 is the typical standard).

For assumption 1, . For assumption 2, .

Because these assumptions do not appear to be violated, it is safe to proceed with a Chi-squared test of independence.

#### Performing Chi-squared Test of Independence

We will perform the test and store the results in `chisq_results`

```{r}
chisq_results <- chisq.test(cont_table, correct = FALSE)

chisq_results
#TODO: prettify the results using janitor (?)
```

sd

## Discussion

After conducting the analysis, we conclude that tree heights vary significantly across neighborhoods. The Chi-squared test of independence result shows the p-value approximately equal to zero thus is less than our predefined significance level of 0.05. Therefore, we can reject null hypothesis that the two categorical variables are independent (there is no association). This means that there is a statistically significant association between neighborhood and tree height.

This reflects differences in tree density in local urban planning. The EDA plot reveals that the RENFREW-COLLINGWOOD neighborhood stands out with the darkest color, indicating the highest concentration of trees, particularly among the shortest height ranges. Other neighborhoods such as VICTORIA-FRASERVIEW, SUNSET, RENFREW-COLLINGWOOD, KENSINGTON-CEDAR COTTAGE, and HASTINGS-SUNRISE tend to have a greater concentration of trees in the lower height ranges (1-4) than other areas. On the other hand, SHAUGHNESSY and DOWNTOWN are notable for having the tallest trees.

Across neighborhoods, the most common tree height range appears to be between 1 and 3 units (height range ID referring to the specific tree height), indicating that the urban tree population is predominantly composed of younger or medium-sized trees. Taller trees (height range 6-10 units) are scarce or entirely absent in most neighborhoods, which may reflect the challenges posed by urban constraints such as limited space, infrastructure interference, or deliberate pruning practices to control growth.

Tree management patterns vary greatly across neighborhoods. Urban areas like DOWNTOWN and FAIRVIEW exhibit consistently light colors across the height ranges, suggesting limited green spaces for tree planting. Surprisingly, even non-urban neighborhoods such as DUNBAR-SOUTHLANDS and ARBUTUS RIDGE show lighter colors, indicating potential land availability that could be utilized for planting new trees. This points to opportunities for the government or community organizations to prioritize these areas for greening initiatives.

Ecologically, neighborhoods with a higher density of trees, such as RENFREW-COLLINGWOOD, enjoy significant environmental benefits, including improved air quality, better urban cooling effects, and enhanced biodiversity. However, the dominance of trees in the 1-3 height range also suggests that many of these trees are young and might require additional care to ensure healthy growth into taller, mature trees.

Note that we do not have the evidence to state these insights and patterns are statstically significant, as the Chi-squared test does not specify which levels are significantly different.

Overall, the findings emphasize the need for targeted greening initiatives, particularly in urban neighborhoods like DOWNTOWN and FAIRVIEW, where tree coverage is limited. Additionally, fostering the growth of taller trees is crucial across all neighborhoods to promote long-term environmental sustainability. Encouraging the planting and care of taller tree species can help balance the urban ecosystem and create more resilient green spaces in the face of growing urbanization.

### 

## References

-   City of Vancouver Open Data Portal: *Street Trees Dataset*. Available at: <https://opendata.vancouver.ca/explore/dataset/street-trees/information/?disjunctive.species_name&disjunctive.common_name&disjunctive.on_street&disjunctive.neighbourhood_name>

-   Wickham, H., & Grolemund, G. (2016). *R for Data Science: Import, Tidy, Transform, Visualize, and Model Data*. O'Reilly Media.

-   Janitor R Package Documentation. Available at: <https://cran.r-project.org/web/packages/janitor/janitor.pdf>

-   ggplot2 R Package Documentation. Available at: <https://ggplot2.tidyverse.org/>

-   Vancouver Urban Forestry Strategy. (2018). City of Vancouver. Available at: <https://vancouver.ca/parks-recreation-culture/urban-forestry-strategy.aspx>
